---
title: "Assignment 1: aberration detection"
author: "Dan Weinberger"
date: "Jan 30, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(surveillance)
library(lubridate)
library(zoo)
library(ciTools)
library(MASS)
library(tidyr)
library(dplyr)
library(ggplot2)
library(shiny)
source('./R/App_Hist_limit.R')
source('./R/FarringtonApp.R')
source('./R/glrpoisApp.R')
source('./R/surv.ds.convert.R')
```

##Part 1 :HIV in Chile

### Goal for the analysis
Your goal is to set up an aberration detection system for hospitalizations associated with HIV in Chile. 

EDAD: age in years
diag1: primary diagnosis, in ICD10 coding
date: Date admitted to hospital 

You have been provided with a subset of the larger database that includes ICD10 codes B20 and Z21, which are codes for HIV/AIDS-related causes of hospitalization: https://icd.who.int/browse10/2016/en#/A00-A09

### First read in the data 

```{r readscsv}
d1 <- read.csv('./Data/chile_hiv.csv')
```

### Tell R that 'date' is a date variable and assign a format
see https://www.statmethods.net/input/dates.html
```{r date_format}
d1$date <- as.Date(d1$date, "%d%b%Y")
```

Then create a new variable called 'week.date' that has the date of the Sunday of the week in which the admission occurred. The lubridate has some functions that make this easier. You want to round the date down to the nearest Sunday. The floor_data function can accomplish this: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf

```{r date_format2}
#"round" the date down to 
d1$week.date <- floor_date(d1$date, unit='week')
```


### Then do some basic explorations. What is the distibution of ages? of Dates? (make a histogram for each)
```{r hist1}
hist(d1$EDAD, xlab='Age (years)' )
```

### Frequency of the codes

Using an online ICD10 dictionary, what are the top 3 diagnosis codes used for HUV/AIDS in this database? 

```{r freq.codes, echo=FALSE}
sort(table(d1$diag1),decreasing=T)
```

Unlike the in-class exercise that we completed, we are not going to extract subsets of codes--we will use all HIV-related codes for our case definition.

```{r}
icd10.3digits<-substr(d1$diag1,1,3) #extract 1st 3 digits from IC10 code
icd10.3digits[1:10] #view first 10
```
### Historgram of ages

```{r}
xx
```


### Let's aggregate now by week.date

call the data frame 'd5', and name your variable with the number of counts 'hiv'
```{r}
XX
```

Plot your time series

```{r}
xx
```



### HIV: Aberration detection algorithms

First try to use historical limits for the weeks in 2007-2008
```{r}
app.hist.limit(ds=d5, datevar='week.date', casevar='hiv')
```



### HIV: Farrington method
```{r hist.limits2}
FarringtonApp(ds=d5, datevar='week.date', casevar='hiv')
```
### HIV CUSUM/GLR method
```{r}
glrpois_App(ds=d5, datevar='week.date', casevar='hiv', n.weeks.train=100)

```


##Part 2: Microcephaly 
Your goal is to set up an aberration detection system for microcephaly in infants in Chile. 

EDAD: age in years
diag1: primary diagnosis, in ICD10 coding
date: Date admitted to hospital 

You have been provided with a subset of the larger database that includes ICD10 codes in the Q chapter, which include congenital malformations: https://www.icd10data.com/ICD10CM/Codes/Q00-Q99

### First read in the data 
Change the directory as needed to point to where the .rds file is saved
```{r read.birth.defects}
e1<-readRDS('./Data/ch.birth.defects.rds')

```

### Tell R that 'date' is a date variable and assign a format
see https://www.statmethods.net/input/dates.html
```{r date_format}
e1$date<- as.Date(e1$date, "%d%b%Y")
```

Then create a new variable called 'week.date' that has the date of the Sunday of the week in which the admission occurred. The lubridate has some functions that make this easier. You want to round the date down to the nearest Sunday. The floor_data function can accomplish this: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf

```{r date_format2}
#"round" the date down to 
e1$week.date<-floor_date(e1$date, unit='week')
```


### Then do some basic explorations. What is the distibution of ages? 
```{r hist1}
par(mfrow=c(1,1))
hist(e1$EDAD, xlab='Age (years)' )
```

### Frequency of the codes

```{r freq.codes, echo=FALSE}
sort(table(e1$diag1), decreasing=T)[1:50]
```


```{r}
icd10.3digits<-substr(e1$diag1,1,3) #extract 1st 3 digits from IC10 code
#icd10.3digits[1:10] #view first 10

#Initialize variables
e1$q02 <- 0
e1$q02[icd10.3digits %in% c('Q02') ]<-1
sum(e1$q02) #check your work
hist(e1$EDAD[e1$q02==1]) #see age distribution for microcephaly

e1$q02[e1$EDAD>0]<-0 #restricts case definition to kids <1 year


```

### Let's aggregate now by week.date

```{r}
e2 <- e1 %>%
  group_by(week.date) %>%
  summarize(q02 = sum(q02)) %>%
  ungroup()
```

It is also important to make sure that the time series is 'filled'--if there are weeks with 0 counts, it needs to be represented in the time series.

```{r}
 e5 <- e2 %>%
  tidyr::complete(week.date=seq.Date(min(week.date, na.rm=T), max(week.date, na.rm=T), 'week'), fill=list(q02=0)) #fills 0s


e5 <- e5[!is.na(e5$week.date),]

e5 <- as.data.frame(e5)
```



```{r}
p2 <- ggplot(e5, aes(x=week.date, y=q02)) +
  geom_line() +
  ylab('Cases of microcephaly') +
  theme_classic()
p2
```


### Microcephaly: Historical limits

```{r}
app.hist.limit(ds=e5, datevar='week.date', casevar='q02')
```

### Microcephaly: Farrington method

Can't use this one with very low counts...

### Microcephaly:GLR/CUSUM method

```{r}
glrpois_App(ds=e5, datevar='week.date', casevar='q02', n.weeks.train=53)

```






